Machine learning enhanced empirical potentials for metals and alloys

### 2021-04-27 doc/papers/ml-eam/manuscript.tex
The revised manuscript for CPC.

\author[a]{Xin Chen}
\author[a]{Li-Fang Wang}
\author[a]{Xing-Yu Gao}
\author[b,a]{Ya-Fan Zhao}
\author[b,a]{De-Ye Lin\corref{author}}
\author[a]{Wei-Dong Chu}
\author[a,b]{Hai-Feng Song\corref{author}}

\cortext[author] {Corresponding author.
\textit{E-mail address:} lindeye0716@163.com, song\_haifeng@iapcm.ac.cn}
\address[a]{Institute of Applied Physics and Computational Mathematics, 
Beijing 100088, China}
\address[b]{CAEP Software Center for High Performance Numerical Simulation, 
Beijing 100088, China}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Abstract
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{abstract}
Interatomic potential (i.e. force-field) plays a vital role in atomistic 
simulation of materials. 
Empirical potentials like the embedded atom method (EAM) and its variant 
angular-dependent potential (ADP) have proven successful in modeling many 
metals qualitatively. In the past few years, machine learning has became a 
compelling approach for modeling interatomic interactions. Powered by big data and 
efficient optimizers, machine learning interatomic potentials can generally 
approximate to the the accuracy of the first-principles calculations based on 
the quantum mechanics theory. In this works, we successfully 
developed a route to express EAM and ADP within machine learning framework in 
highly-vectorizable form and further incorperated several physical constraints 
into the training. As it is proved in this work, the performances of empirical 
potentials can be significantly boosted with few traning data.
For energy and force 
predictions, machine lear̦ned EAM and ADP can be almost as accurate as the 
spectral neighbor analysis potential (SNAP) on the fcc Ni, bcc Mo and Mo-Ni 
alloy systems, while the molecular dynamics simulations performed by our 
potentials are nearly 1000 times faster. Machine learned EAM and ADP can also 
reproduce some key materials properties, such as elastic constants, melting 
temperatures and surface energies, close to the first-principles accuracy. 
Our results suggest a new and systematic route for developing machine learning 
interatomic potentials. All the new algorithms have been implemented in our 
program TensorAlloy.
\end{abstract}

@@

{\bf PROGRAM SUMMARY} \\
\begin{small}
\noindent
{\em Program Title: TensorAlloy} \\
{\em Licensing provisions(please choose one): LGPL}             \\
{\em Programming language: Python 3.7}                          \\
{\em Supplementary material:}                                   \\
{\em Journal reference of previous version: 
https://doi.org/10.1016/j.cpc.2019.107057}                      \\
{\em Does the new version supersede the previous version?:} YES \\
{\em Reasons for the new version: This new version is a significant extension to 
the previous version. Now machine learning approaches and physical constraints 
can be used together to tune empirical potentials (for example the embbeded atom 
method). Machine learning optimized empirical potentials can be almost as 
accurate as machine learning interaction potentials but run much faster} \\
{\em Summary of revisions:}*\\
{\em Nature of problem:} \\
  Optimizing empirical potentials with machine learning approaches and physical 
  constraints. \\
{\em Solution method:} \\
  The TensorAlloy program is built upon TensorFlow and the virtual-atom 
  approach. we successfully developed a route to express the embbeded atom 
  method and the angular-dependent potential within machine learning framework 
  in highly-vectorizable form and further enhanced the potentials with physical
  constraints. Machine learning can significantly boost their performances with 
  few training data.   \\
{\em Additional comments including Restrictions and Unusual features}\\
  This program needs TensorFlow 1.14.*. Neither newer or older TensorFlow is 
  supported. \\
\end{small}

@@

Atomistic modeling plays a vital role in materials science. \textit{Ab initio} 
calculation or force-field based molecular dynamics simulation (MD) are 
effective ways to study, understand or predict chemical and physical properties 
of materials. \textit{ab initio} approaches are generally much more precise but 
they are rarely used on large-scale metallic systems due to their extremely-high 
computation expenses. Physical model based empirical potential (force-field), 
such as the embedded-atom method (EAM) \cite{EAM0,EAM1,EAM2,EAM3,EAM4}, 
modified embedded-atom method (MEAM) \cite{MEAM0,MEAM1,MEAM2,MEAM_Jelinek}, 
bond-order potential (BoP) \cite{BOP0,BOP1,BOP2,BOP3}, or angular-dependent 
potential (ADP) \cite{ADP0,ADP1,ADP2,ADP3,ADP_FeNi}, still plays the major role 
in long-time simulations and these empirical methods can achieve reasonable 
accuracy with much lower computation costs. Empirical potentials 
generally have very few learnable parameters and both microscopic observables 
(energy, forces, virial, etc.) and macroscopic observables (melting 
point, surface energy, etc.) can be used to tune these parameters. Finding 
optimal parameters of empirical potentials is always a challenging task. Global 
optimization (GO) approaches (Basin-Hopping \cite{BH0,BH1}, genetic algorithm 
\cite{PCIPY,PGA}, etc) are traditionally used to find the best possible 
parameters. However, the gradients of the losses with respect to model 
parameters are difficult or even impossible to calcualte. Hence, 
global optimizations are generally not that effective.

In the last decade, machine learning (ML) has become one of the hottest 
topics in many research areas. In the materials science, researchers have made 
great efforts on developing ML models to describe atomic interactions. Such ML 
models are considered as machine learning interaction potentials (MLIPs). Until 
now, hundreds of MLIPs have been proposed. Among them, the symmetry-function 
based atomistic neural network (ANN) model, published by Parinello and Behler in 
2007 \cite{SF2007,SF1,SF2,SF3,SF4}, is still the most popular choice in modeling 
metallic interactions \cite{SFApp0,SFApp1,SFApp2}. 
The smooth overlap atomic positions descriptor based gaussian approximation 
potential (SOAP-GAP) \cite{soap0,soap1,soap2,soap3}, developed by Bartòk et al, 
can give extremely accurate prediction results, although it's a bit 
computationally expensive. 
Recently, Thompson and co-workers proposed another quantum-accurate MLIP named 
the spectral neighbor analysis potential (SNAP) \cite{SNAP_Algo} and it has 
been proven working on a broad range of metals and 
alloys \cite{SNAP_Mo,SNAP,SNAP_New}.

In many cases, MILPs can easily outperform state-of-art empirical potentials. 
Compared with empirical potentials, MLIPs generally have orders of magnitudes 
more model paramters. The redundant parameter space greatly reduces the 
difficulty of fitting complicated potential energy surfaces. But, to effectively 
train a MLIP and avoid overfitting, a large high-quality (versatile) training 
dataset is probably needed. However, MILPs \textbf{can really} take advantages 
of "big data" for two reasons. First, MLIPs typically only have 
\textit{basic} or \textit{simple} arithmetic operations. Thus, MILPs can be 
implemented within modern deep learning frameworks 
(TensorFlow \cite{tensorflow}, PyTorch \cite{pytorch}, etc) 
so that the gradients of the total loss with respect to fitting parameters can 
be can obtained with the backpropagation algorithm automatically and 
efficiently. Second, MLIPs are mostly vectorizable. Hence, GPUs can be utilized
to significantly accelerate training and using of MILPs.

However, MLIPs also have challenges. The large parameter space and lack of 
physical background makes the "big data" a requisite. The cost of dataset is 
non-negligible. Besides, even "big data" can only cover a small portion of real 
physical environments (temperature, external pressure, etc). Outside the 
training zone, the performances of MILPs may not that stable. For long-time 
molecular dynamics (MD) simulations of large-scale ($10^5$ or more) systems, 
computation efficiency also becomes a major concern. Recent benchmark 
tests \cite{SnapBenchmark} suggest that MILPs are still too expensive. 
At present, most MLIPs are used to examine small to medium ($10^3$ to $10^4$) 
systems \cite{SNAP,SFApp0,SFApp1,SFApp2,soap3}.

In this work, instead of designing new atomic descriptors, we chose a new route 
to develop MLIP: combining machine learning with empirical potentials. We 
successfully implemented EAM and its variant ADP within TensorFlow so that 
machine learning approaches can be used directly to tune EAM and ADP potentials. 
Our results suggest ML-EAM or ML-ADP can be as precise as the SNAP machine 
learning method.

@@

% -----------
% Section 2.B Transformation
% -----------
\subsection{Transformation}
\label{sec:transformation}

@@

To find optimal parameters of the funcitons in section \ref{sec:functions}, the 
overall loss function shall be defined. Just like other machine learning tasks, 
the mini-batch stochastic gradient descent algorithm is used to minimize the 
loss function. 
The Adam optimizer \cite{adam} is used to minimize Equation \ref{eq:loss}. 
In most cases, we use 0.01 as the initial learning rate and the batch size 
ranges from 20 to 50. Since there are very few adjustable parameters compared 
with neural network potentials, the optimization typically needs very few 
training steps to converge. When the optimization is finished, the corresponding 
Lammps setfl potential file will be exported.

@@

The original Zjw04 potentials were fitted to experimentally obtained quantities.
Thus, we use the relative MAE:
\begin{equation}
\mathbf{rMAE} = \frac{1}{N}\sum_{i=1}^{N}{
    | \frac{(E_i - E_i^{dft})}{N_i} - (E_{eq} - E_{eq}^{dft}) |
}
\end{equation}
to measure their performaces. Here $N_i$ is the number of atoms in structure 
$i$, $E_i$ is the Zjw04 energy of structure $i$ while $E_i^{dft}$ represents its 
corresponding DFT energy, $E_{eq}$ and $E_{eq}^{dft}$ refers to the equilibrium 
energy of Zjw04 and DFT, respectively.

@@

The above results suggest that with machine learning and essential physical 
constraints, for the fcc and bcc solids, the traditional empirical potentials, 
EAM/ADP, can be improved to be almost as accurate as the SNAP method using 
exactly the same dataset. But, EAM/ADP are around $10^2$ to $10^3$ times faster 
than SNAP. To understand this, we can start from the many-body expansion (MBE) 
scheme \cite{kCON}.

@@

This work was supported by the National Key Research and Development Program of 
China under Grant No. 2016YFB0201204, the Science Challenge Project under Grant 
No. TZ2018002 and the National Natural Science Foundation of China under Grant 
No. U1630250.

>>>>

\author[a]{Xin Chen}
\author[a]{Li-Fang Wang}
\author[a]{Xing-Yu Gao}
\author[b,c]{Ya-Fan Zhao}
\author[b,c]{De-Ye Lin\corref{author}}
\author[a]{Wei-Dong Chu}
\author[a]{Hai-Feng Song\corref{author}}

\cortext[author] {Corresponding author.
\textit{E-mail address:} lindeye0716@163.com, song\_haifeng@iapcm.ac.cn}
\address[a]{Laboratory of Computational Physics, Institute of Applied Physics 
and Computational Mathematics, Beijing 100088, China}
\address[b]{CAEP Software Center for High Performance Numerical Simulation, 
Beijing 100088, China}
\address[c]{Institute of Applied Physics and Computational Mathematics, Beijing 
100088, China}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Abstract
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{abstract}
Interatomic potential (i.e. force-field) plays a vital role in atomistic 
simulation of materials. 
Empirical potentials like the embedded atom method (EAM) and its variant 
angular-dependent potential (ADP) have proven successful in many 
metals. In the past few years, machine learning has became a 
compelling approach for modeling interatomic interactions. Powered by big data 
and efficient optimizers, machine learning interatomic potentials can generally 
approximate to the the accuracy of the first-principles calculations based on 
the quantum mechanics theory. In this works, we successfully 
developed a route to express EAM and ADP within machine learning framework in 
highly-vectorizable form and further incorperated several physical constraints 
into the training. As it is proved in this work, the performances of empirical 
potentials can be significantly boosted with few traning data.
For energy and force 
predictions, machine tuned EAM and ADP, can be 
almost as accurate as the computationally expensive spectral neighbor analysis 
potential (SNAP) on the fcc Ni, bcc Mo and Mo-Ni alloy systems. 
Machine learned EAM and ADP can also 
reproduce some key materials properties, such as elastic constants, melting 
temperatures and surface energies, close to the first-principles accuracy. 
Our results suggest a new and systematic route for developing machine learning 
interatomic potentials. All the new algorithms have been implemented in our 
program TensorAlloy.
\end{abstract}

@@

{\bf PROGRAM SUMMARY} \\
\begin{small}
\noindent
{\em Program Title: TensorAlloy} \\
{\em Licensing provisions(please choose one): LGPL}             \\
{\em Programming language: Python 3.7}                          \\
{\em Supplementary material:}                                   \\
{\em Journal reference of previous version: 
https://doi.org/10.1016/j.cpc.2019.107057}                      \\
{\em Does the new version supersede the previous version?:} YES \\
{\em Reasons for the new version: This new version is a significant extension to 
the previous version. Now machine learning approaches and physical constraints 
can be used together to tune empirical potentials (for example the embbeded atom 
method). Machine learning optimized empirical potentials can be almost as 
accurate as machine learning interaction potentials but run much faster.} \\
{\em Summary of revisions:}*\\
{\em Nature of problem:} \\
  Optimizing empirical potentials with machine learning approaches and physical 
  constraints. \\
{\em Solution method:} \\
  The TensorAlloy program is built upon TensorFlow and the virtual-atom 
  approach. we successfully developed a route to express the embbeded atom 
  method and the angular-dependent potential within machine learning framework 
  in highly-vectorizable form and further enhanced the potentials with physical
  constraints. Machine learning can significantly boost their performances with 
  few training data.   \\
{\em Additional comments including Restrictions and Unusual features}\\
  This program needs TensorFlow 1.14.*. Neither newer or older TensorFlow is 
  supported. \\
\end{small}

@@

Atomistic modeling plays a vital role in materials science. \textit{Ab initio} 
calculation or force-field based molecular dynamics simulation (MD) are 
effective ways to study, understand or predict chemical and physical properties 
of materials. \textit{ab initio} approaches are generally much more precise but 
they are rarely used on large-scale metallic systems due to their extremely-high 
computational expenses. Physical model based empirical potentials (force-field), 
such as the embedded-atom method (EAM) \cite{EAM0,EAM1,EAM2,EAM3,EAM4}, 
modified embedded-atom method (MEAM) \cite{MEAM0,MEAM1,MEAM2,MEAM_Jelinek}, 
bond-order potential (BoP) \cite{BOP0,BOP1,BOP2,BOP3}, or angular-dependent 
potential (ADP) \cite{ADP0,ADP1,ADP2,ADP3,ADP_FeNi}, still play the major role 
in long-time simulations and these empirical methods can achieve reasonable 
accuracy with much lower computation costs. Empirical potentials 
generally have very few learnable parameters and both microscopic observables 
(energy, forces, virial, etc.) and macroscopic observables (melting 
point, surface energy, etc.) can be used to tune these parameters. Finding 
optimal parameters of empirical potentials is always a challenging task. Global 
optimization (GO) approaches (Basin-Hopping \cite{BH0,BH1}, genetic algorithm 
\cite{PCIPY,PGA}, etc) are traditionally used to find the best possible 
parameters. However, the gradients of the losses with respect to model 
parameters are difficult or even impossible to calculate. Hence, 
global optimizations are generally not that effective.

In the last decade, machine learning (ML) has become one of the hottest 
topics in many research areas. In the field of materials science, researchers 
have made great efforts on developing ML models to describe atomic interactions. 
Such ML 
models are considered as machine learning interaction potentials (MLIPs). Until 
now, hundreds of MLIPs have been proposed. Among them, the symmetry-function 
based atomistic neural network (ANN) model, published by Parinello and Behler in 
2007 \cite{SF2007,SF1,SF2,SF3,SF4}, is still the most popular choice in modeling 
metallic interactions \cite{SFApp0,SFApp1,SFApp2}. 
The smooth overlap atomic positions descriptor based gaussian approximation 
potential (SOAP-GAP) \cite{soap0,soap1,soap2,soap3}, developed by Bartòk et al, 
can give extremely accurate prediction results, although it's a bit 
computationally expensive. 
Recently, Thompson and co-workers proposed another quantum-accurate MLIP named 
the spectral neighbor analysis potential (SNAP) \cite{SNAP_Algo} and it has 
been proven working on a broad range of metals and 
alloys \cite{SNAP_Mo,SNAP,SNAP_New}.

In many cases, MILPs can easily outperform state-of-art empirical potentials. 
Compared with empirical potentials, MLIPs generally have orders of magnitudes 
more model paramters. The redundant parameter space greatly reduces the 
difficulty of fitting complicated potential energy surfaces. But, to effectively 
train a MLIP and avoid overfitting, a large high-quality (versatile) training 
dataset is probably needed. However, MILPs \textbf{can really} take advantages 
of "big data" for two reasons. First, MLIPs typically only have 
\textit{basic} or \textit{simple} arithmetic operations. Thus, MILPs can be 
implemented within modern deep learning frameworks 
(TensorFlow \cite{tensorflow}, PyTorch \cite{pytorch}, etc) 
so that the gradients of the total loss with respect to fitting parameters can 
be can obtained with the backpropagation algorithm automatically and 
efficiently. Second, MLIPs are mostly vectorizable. Hence, graphic processing 
units (GPUs) can be utilized to significantly accelerate training and using of 
MILPs.

However, MLIPs also have challenges. The large parameter space and lack of 
physical background makes the "big data" a requisite. The cost of dataset is 
non-negligible. Besides, even "big data" can only cover a small portion of real 
physical environments (temperature, external pressure, etc). Outside the 
training zone, the performances of MILPs may not that stable. For long-time 
molecular dynamics (MD) simulations of large-scale ($10^5$ or more) atoms, 
computation efficiency also becomes a major concern. Recent benchmark 
tests \cite{SnapBenchmark} suggest that MILPs are still too expensive. 
At present, most MLIPs are used to examine small to medium ($10^3$ to $10^4$) 
atoms \cite{SNAP,SFApp0,SFApp1,SFApp2,soap3}.

In this work, instead of designing new atomic descriptors, we chose a new route 
to develop MLIP: combining machine learning approach with empirical potentials. 
We successfully implemented EAM and its variant ADP within TensorFlow so that 
machine learning approaches can be used directly to tune EAM and ADP potentials. 
Our results suggest ML-EAM or ML-ADP can be as precise as the SNAP machine 
learning method.

@@

% -----------
% Section 2.B Transformation
% -----------
\subsection{Vectorizable Transformation}
\label{sec:transformation}

@@

The Adam optimizer \cite{adam} is used to minimize Equation \ref{eq:loss}. 
In most cases, we use 0.01 as the initial learning rate and the batch size 
ranges from 20 to 50. Since there are very few adjustable parameters compared 
with neural network potentials, the optimization typically needs very few 
training steps to converge. When the optimization is finished, the corresponding 
Lammps setfl potential file will be exported.

The number of trainable parameters in empirical potentials are quite limited. 
According to our tests, typically tens of thousands of steps are sufficient to 
finish a training. The training time mainly depends on the choices of losses. 
Training with standard metrics (energy, forces and stress) only needs a few GPU 
hours. Including the EOS constraint will just slightly increase the time cost.
However, the elastic constraint is extremely complicated. Adding elastic term 
to the total loss will significantly increase the training time to 
approximately 20 to 40 GPU hours depending on your hardware and the number of 
elastic losses. In our experience, training with energy, forces, stress and EOS 
constraints first then restarting from a checkpoint with elastic constraints 
appended is the best strategy.

@@

The original Zjw04 potentials were fitted to experimental data. It would be 
unfair if we calculate their relative errors based on DFT data directly. Hence, 
in this work we also use the relative MAE (rMAE), 
followed by Shyue Ping Ong\cite{SNAP,SNAP_Mo}:
\begin{equation}
\mathbf{rMAE} = \frac{1}{N}\sum_{i=1}^{N}{
    | \frac{(E_i - E_i^{dft})}{N_i} - (E_{eq} - E_{eq}^{dft}) |
}
\end{equation}
to measure their performaces. Here $N_i$ is the number of atoms in structure 
$i$, $E_i$ is the Zjw04 energy of structure $i$ while $E_i^{dft}$ represents its 
corresponding DFT energy, $E_{eq}$ and $E_{eq}^{dft}$ refers to the equilibrium 
energy of Zjw04 and DFT, respectively.

@@

The above results suggest that with machine learning and essential physical 
constraints, for the fcc and bcc solids, the traditional empirical potentials, 
EAM/ADP, can be improved to be almost as accurate as the SNAP method using 
exactly the same dataset. But, EAM/ADP are around $10^2$ to $10^3$ times faster 
than SNAP, according to the benchmark tests did by Trott and Thompson, who are 
founders of SNAP\cite{SnapBenchmark}. To understand this, we can start from 
the many-body expansion (MBE) scheme \cite{kCON}.

@@

This work was supported by the National Key Research and Development Program of 
China under Grant No. 2016YFB0201204, the Science Challenge Project under Grant 
No. TZ2018002, the National Natural Science Foundation of China under Grant 
No. U1630250, No. 12002064 and No. 12004046.
